---
title: "PACMOS-MESOMICS-fuzzy-tutorial"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
self_contained: true
vignette: >
  %\VignetteIndexEntry{1. PACMOS-MESOMICS-fuzzy-tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css, echo=FALSE}
pre,
pre.output,
pre.message,
pre.warning,
pre.sourceCode,
div.sourceCode {
  max-height: 250px;
  overflow-y: auto;
  overflow-x: auto;
}

```

# Introduction

This vignette shows how to:

0. Installation of PACMOS.
1. Add a query sample to MOFA reference inputs,
2. Run MOFA2,
3. Project query sample into reference MOFA space,
4. Infer archetype weights of query sample,
5. Plot query sample onto Pareto archetype space.

## Step 0: Installation

Before using MESOMICS, make sure you have the required Python environment.

For Linux/macOS, open a terminal.  

```{bash, eval=FALSE}
git clone https://github.com/IARCbioinfo/PACMOS.git
cd PACMOS
conda env create -f pacmos_env.yml
conda activate pacmos_env

# Linux/macOS users:
export PACMOS_PYTHON=$(which python)
echo $PACMOS_PYTHON
```

For Windows, open the **Anaconda Prompt** and run:
```{bash, eval=FALSE}
conda activate pacmos_env
where python
```

This will return a path similar to:

```text
C:\Users\YOUR_USERNAME\miniconda3\envs\pacmos_env\python.exe
```

Then in R run:

```{r, eval=FALSE}
Sys.setenv(
  PACMOS_PYTHON =
    "C:/Users/YOUR_USERNAME/miniconda3/envs/pacmos_env/python.exe"
)
```

```{r install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("IARCbioinfo/PACMOS", dependencies = TRUE)
```

```{r setup}
library(PACMOS)
```

Example data paths
```{r setup-paths}
mofa_dir <- system.file("extdata/MESOMICS_references", package = "PACMOS")
reference_LFs  <- system.file("extdata/MESOMICS_references", "MESOMICS_latent_factors.csv", package = "PACMOS")
query_csv <- system.file("extdata/test_data", "MESOMICS_test_expr.csv", package = "PACMOS")


out_dir <- file.path('.', "MESOMICS_output")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
out_dir

python_path <- Sys.getenv("PACMOS_PYTHON")

if (!file.exists(python_path)) {
  stop(
    "PACMOS_PYTHON is not set correctly.\n",
    "Activate pacmos_env and set PACMOS_PYTHON to your conda python path."
  )
}
```

## Step 1: Add query samples to MOFA inputs
This function integrates one or more query samples into reference multi-omic MOFA input matrices using an incremental approach. For each query sample, it aligns gene identifiers with the reference matrices, inserts observed values into the corresponding data layer, and fills all other layers with NA to maintain a consistent structure across views.

The function creates a structured output directory where each query sample has its own folder, containing an inputs/ subdirectory with all .RData matrices required for downstream MOFA training.

```{r run-step1, results='markup', warning=FALSE}

## -----------------------------------------------------------------------------
print('Starting Step 1')
PACMOS::s1_add_sample_to_mofa(
  query_matrix_path = query_csv,
  mofa_dir = mofa_dir,
  value_data_types = c('D_exprB_MOFA'),
  outdir = out_dir,
  python_bin = python_path
)
cat("Outputs are saved in:", out_dir, "\n")

```

## Step 2: Train MOFA models
This function trains MOFA models for each query sample using the prepared input matrices. It loads the multi-omic inputs generated in Step 1, configures the model according to the specified views and data types (including binary layers), and runs MOFA.

Each query sample is processed independently, producing a trained MOFA model.

```{r run-step2, warning=FALSE}
print('Starting Step 2')

PACMOS::s2_run_mofa(
  models_dir      = out_dir,
  matrices_subdir = "inputs",
  num_factors = 10,
  convergence_mode = "slow",
  maxiter = 10000,
  use_basilisk = FALSE,
  skip_existing = TRUE,
  python_bin      = python_path,
  views_map = c(
    RNA     = "D_exprB_MOFA",
    MethPro = "D_met.proB_MOFA",
    MethBod = "D_met.bodB_MOFA",
    MethEnh = "D_met.enhB_MOFA",
    Total   = "D_cnv_MOFA",
    Minor   = "D_loh_MOFA",
    Alt     = "D_alt_MOFA"
  ),
  binary_views = "Alt"
)
cat("Outputs are saved in:", out_dir, "\n")

```

## Step 3: Visualize query samples
This function projects query samples into the reference/retrained latent factor space and generates quality control visualizations. It compares the learned latent factors of retrained models with reference axes, computes correlations, and produces diagnostic plots to assess how well the query samples align with known biological structure.

The outputs are saved as PDF files for each sample in the `outputs` directory inside each query sample folder.

```{r run-step3,warning=FALSE}
print('Starting Step 3')

sample_dirs <- list.dirs(out_dir, recursive = FALSE, full.names = TRUE)
sample_dirs <- sample_dirs[dir.exists(file.path(sample_dirs, "inputs"))]
sample_ids <- basename(sample_dirs)

s3_results <- list()

for (id in sample_ids) {
  s3_results[[id]] <- PACMOS::s3_plot_query_samples_mofa(
    models_dir      = out_dir,
    matrices_subdir = "inputs",
    query_sample    = id,
    reference_LFs   = reference_LFs,
    reference_axes  = c("Morphology_LF", "Adaptive-response_LF"),
    group           = "group1",
    python_bin      = python_path
  )
}
cat("Outputs are saved in:", out_dir, "\n")

```

### Interpreting Step 3 quality control

Step 3 checks whether the latent factors learned after adding the query sample can be aligned back to the original reference MOFA space.

The QC PDF contains three main diagnostics (Good alignment is indicated by high absolute correlation, a clear diagonal trend in the scatter plot, and relatively low RMSE):

1. **Factor assignment summary**: shows which retrained MOFA factor was matched to each reference latent factor.

2. **Correlation heatmap**: The heatmap shows correlations between retrained MOFA factors and reference latent factors. Rows are retrained MOFA factors and columns are reference axes. The black boundary marks the selected factor for each reference axis.

```{r step3-qc-heatmap, fig.width=7, fig.height=5}
sample_to_show <- sample_ids[1]

s3_results[[sample_to_show]]$plots$alignment_heatmap
```

3. **Scatter plots**: The scatter plot compares reference LF values with aligned retrained LF values. A strong diagonal trend and high correlation indicate successful alignment.
```{r step3-qc-scatter, fig.width=6, fig.height=5}
s3_results[[sample_to_show]]$plots$scatter_by_axis$Morphology_LF
```


The projection plot shows the query sample highlighted in red within the reference latent factor space (grey points).
```{r step3-projection, fig.width=6, fig.height=5}
s3_results[[sample_to_show]]$plots$projection_pairs[[1]]
```

## Step 4: Infer archetype weights
This function estimates the contribution of predefined biological archetypes to each query sample using a fuzzy weighting approach. Based on the latent factor coordinates obtained from MOFA, it calculates the degree to which each sample belongs to each archetype.

The result is saved as a csv file in the `outputs` directory.

```{r run-step4, warning=FALSE}
print('Starting Step 4')

archetype_coords <- data.frame(
  Archetype = c("Cell division", "Tumor-immune-interaction", "Acinar"),
  Morphology_factor.MESOMICS = c(-3.59733245242571, -1.79807036681759, 3.85856701073313),
  Adaptive.response_factor.MESOMICS = c(-2.39600999565261, 3.56348237698459, -0.825586204203109),
  stringsAsFactors = FALSE)

PACMOS::infer_fuzzy_weights(
  models_dir      = out_dir,
  coord           = archetype_coords,
  n_archetypes    = 3,
  input_type      = "stable"
)
cat("Outputs are saved in:", out_dir, "\n")

```

## Step 5: Plot results
This function visualizes the archetype composition of query samples based on the inferred fuzzy weights.

```{r run-step5, warning=FALSE}
print('Starting Step 5')

fuzzy_results <- PACMOS::plot_fuzzy_query_sample(
  models_dir      = out_dir
)
cat("Outputs are saved in:", out_dir, "\n")

```

```{r s5-ternary, fig.width=20, fig.height=80,out.width="700px"}
sample_to_show <- sample_ids[1]

knitr::include_graphics(
  fuzzy_results$samples[[sample_to_show]]$files$projection_pdf
)
```

## Session information

```{r session-info}
sessionInfo()
```
