---
title: "PACMOS-MESOMICS-fuzzy-tutorial"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
self_contained: true
vignette: >
  %\VignetteIndexEntry{1. PACMOS-MESOMICS-fuzzy-tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css, echo=FALSE}
pre,
pre.output,
pre.message,
pre.warning,
pre.sourceCode,
div.sourceCode {
  max-height: 250px;
  overflow-y: auto;
  overflow-x: auto;
}

```

# Introduction
PACMOS is an R package designed to project new query samples into a reference multi-omics latent space derived from MOFA models, enabling sample-level interpretation within an established biological context. It is particularly suited for translational and cancer multi-omics applications where robust comparison of new samples against established molecular reference cohorts is required.

This vignette demonstrates the PACMOS workflow using the  pleural mesothelioma tumours dataset from the MESOMICS cohort, based on the integrative multi-omics framework described at https://doi.org/10.1038/s41588-023-01321-1.

The reference datasets required by PACMOS are provided within the package under inst/extdata/MESOMICS_references/. These reference objects consist of MOFA-compatible matrices where rows correspond to genes or molecular features and columns correspond to reference samples across multiple omics layers.

Query samples are expected as one or more CSV files in which the first column contains gene or feature identifiers, followed by one or more columns representing query samples. Query datasets should use feature identifiers compatible with the reference matrices and may contain one or several omics modalities corresponding to the reference MOFA views.

This vignette demonstrates how to:

0. Install PACMOS
1. Add query samples to MOFA reference inputs
2. Run MOFA2 retraining
3. Project query samples into the reference MOFA latent space
4. Infer archetype weights of query samples
5. Visualize query samples within Pareto archetype space

## Step 0: Installation

Before using PACMOS, make sure you have the required Python environment.

For Linux/macOS, open a terminal.  

```{bash, eval=FALSE}
git clone https://github.com/IARCbioinfo/PACMOS.git
cd PACMOS
conda env create -f pacmos_env.yml
conda activate pacmos_env

# Linux/macOS users:
export PACMOS_PYTHON=$(which python)
echo $PACMOS_PYTHON
```

For Windows, download the PACMOS repository as a ZIP file from https://github.com/IARCbioinfo/PACMOS. Extract the ZIP archive, then open the **Anaconda Prompt** and run:

```{bash, eval=FALSE}
cd <dir-to-PACMOS>\PACMOS-main

dir

conda env create -f pacmos_env.yml
conda activate pacmos_env

where python
```

This will return a path similar to:


```text
## Linux/macOS
/home/user/miniconda3/envs/pacmos_env/bin/python

## windows
C:/Users/YOUR_USERNAME/miniconda3/envs/pacmos_env/python.exe```
```

Before installing PACMOS with vignettes, set `PACMOS_PYTHON` in your R console to the Python executable from your own `pacmos_env` Conda environment. Replace the example path below with your local path.
In R:

```{r, eval=TRUE}
## Linux/macOS
Sys.setenv(
  PACMOS_PYTHON =
    "/home/lipikal/miniconda3/envs/pacmos_env/bin/python" # replace with your path
)

## Windows
#Sys.setenv(
#  PACMOS_PYTHON =
#    "C:/Users/YOUR_USERNAME/miniconda3/envs/pacmos_env/python.exe"
#)
```

```{r install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("IARCbioinfo/PACMOS", dependencies = TRUE)
```

```{r setup}
library(PACMOS)
```

Example data paths
```{r setup-paths}
mofa_dir <- system.file("extdata/MESOMICS_references", package = "PACMOS")
reference_LFs  <- system.file("extdata/MESOMICS_references", "MESOMICS_latent_factors.csv", package = "PACMOS")
query_csv <- system.file("extdata/test_data", "MESOMICS_test_expr.csv", package = "PACMOS")


out_dir <- file.path('.', "MESOMICS_output")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
out_dir

python_path <- Sys.getenv("PACMOS_PYTHON", unset = "")

run_pacmos <- nzchar(python_path) && file.exists(python_path)

if (!run_pacmos) {
  stop(
    "PACMOS_PYTHON is not set to a valid Python path.\n",
    "Before building this vignette, create and activate the PACMOS Conda ",
    "environment, then set PACMOS_PYTHON to its Python executable."
  )
}
```

## Step 1: Add query samples to MOFA inputs
This function integrates one or more query samples into reference multi-omic MOFA input matrices using an incremental approach. For each query sample, it aligns gene identifiers with the reference matrices, inserts observed values into the corresponding data layer, and fills all other layers with NA to maintain a consistent structure across views.

The function creates a structured output directory where each query sample has its own folder, containing an inputs/ subdirectory with all .RData matrices required for downstream MOFA training.

```{r run-step1, results='markup', warning=FALSE}

## -----------------------------------------------------------------------------
print('Starting Step 1')
PACMOS::s1_add_sample_to_mofa(
  query_matrix_path = query_csv,
  mofa_dir = mofa_dir,
  value_data_types = c('D_exprB_MOFA'),
  outdir = out_dir,
  python_bin = python_path
)
cat("Outputs are saved in:", out_dir, "\n")

```

## Step 2: Train MOFA models
This function trains MOFA models for each query sample using the prepared input matrices. It loads the multi-omic inputs generated in Step 1, configures the model according to the specified views and data types (including binary layers), and runs MOFA.

Each query sample is processed independently, producing a trained MOFA model.

```{r run-step2, warning=FALSE}
print('Starting Step 2')

PACMOS::s2_run_mofa(
  models_dir      = out_dir,
  matrices_subdir = "inputs",
  num_factors = 10,
  convergence_mode = "slow",
  maxiter = 10000,
  use_basilisk = FALSE,
  skip_existing = TRUE,
  python_bin      = python_path,
  views_map = c(
    RNA     = "D_exprB_MOFA",
    MethPro = "D_met.proB_MOFA",
    MethBod = "D_met.bodB_MOFA",
    MethEnh = "D_met.enhB_MOFA",
    Total   = "D_cnv_MOFA",
    Minor   = "D_loh_MOFA",
    Alt     = "D_alt_MOFA"
  ),
  binary_views = "Alt"
)
cat("Outputs are saved in:", out_dir, "\n")

```

## Step 3: Visualize query samples
This function projects query samples into the reference/retrained latent factor space and generates quality control visualizations. It compares the learned latent factors of retrained models with reference axes, computes correlations, and produces diagnostic plots to assess how well the query samples align with known biological structure.

The outputs are saved as PDF files for each sample in the `outputs` directory inside each query sample folder.

```{r run-step3,warning=FALSE}
print('Starting Step 3')

sample_dirs <- list.dirs(out_dir, recursive = FALSE, full.names = TRUE)
sample_dirs <- sample_dirs[dir.exists(file.path(sample_dirs, "inputs"))]
sample_ids <- basename(sample_dirs)

s3_results <- list()

for (id in sample_ids) {
  s3_results[[id]] <- PACMOS::s3_plot_query_samples_mofa(
    models_dir      = out_dir,
    matrices_subdir = "inputs",
    query_sample    = id,
    reference_LFs   = reference_LFs,
    reference_axes  = c("Morphology_LF", "Adaptive-response_LF"),
    group           = "group1",
    python_bin      = python_path
  )
}
cat("Outputs are saved in:", out_dir, "\n")

```

### Interpreting Step 3 quality control

Step 3 checks whether the latent factors learned after adding the query sample can be aligned back to the original reference MOFA space.

The QC PDF contains three main diagnostics (Good alignment is indicated by high absolute correlation, a clear diagonal trend in the scatter plot, and relatively low RMSE):

1. **Factor assignment summary**: shows which retrained MOFA factor was matched to each reference latent factor.

2. **Correlation heatmap**: The heatmap shows correlations between retrained MOFA factors and reference latent factors. Rows are retrained MOFA factors and columns are reference axes. The black boundary marks the selected factor for each reference axis.

```{r step3-qc-heatmap, fig.width=7, fig.height=5}
sample_to_show <- sample_ids[1]

s3_results[[sample_to_show]]$plots$alignment_heatmap
```

3. **Scatter plots**: The scatter plot compares reference LF values with aligned retrained LF values. A strong diagonal trend and high correlation indicate successful alignment.
```{r step3-qc-scatter, fig.width=6, fig.height=5}
s3_results[[sample_to_show]]$plots$scatter_by_axis$Morphology_LF
```


The projection plot shows the query sample highlighted in red within the reference latent factor space (grey points).
```{r step3-projection, fig.width=6, fig.height=5}
s3_results[[sample_to_show]]$plots$projection_pairs[[1]]
```

## Step 4: Infer archetype weights
This function estimates the contribution of predefined biological archetypes to each query sample using a fuzzy weighting approach. Based on the latent factor coordinates obtained from MOFA, it calculates the degree to which each sample belongs to each archetype.

The result is saved as a csv file in the `outputs` directory.

```{r run-step4, warning=FALSE}
print('Starting Step 4')

archetype_coords <- data.frame(
  Archetype = c("Cell division", "Tumor-immune-interaction", "Acinar"),
  Morphology_LF = c(-3.59733245242571, -1.79807036681759, 3.85856701073313),
  `Adaptive-response_LF` = c(-2.39600999565261, 3.56348237698459, -0.825586204203109),
  stringsAsFactors = FALSE,
  check.names = FALSE)

PACMOS::infer_fuzzy_weights(
  models_dir      = out_dir,
  coord           = archetype_coords,
  n_archetypes    = 3,
  input_type      = "stable",
  reference_axes = c("Morphology_LF", "Adaptive-response_LF")
)
cat("Outputs are saved in:", out_dir, "\n")

```

## Step 5: Plot results
This function visualizes the archetype composition of query samples based on the inferred fuzzy weights.

```{r run-step5, warning=FALSE}
print('Starting Step 5')

fuzzy_results <- PACMOS::plot_fuzzy_query_sample(
  models_dir      = out_dir
)
cat("Outputs are saved in:", out_dir, "\n")

```

```{r s5-ternary, fig.width=20, fig.height=80,out.width="700px"}
sample_to_show <- sample_ids[1]

knitr::include_graphics(
  fuzzy_results$samples[[sample_to_show]]$files$projection_pdf
)
```

## Session information

```{r session-info}
sessionInfo()
```
