---
title: "PACMOS-tutorial"
output:
  rmarkdown::html_vignette:
    toc: true
    self_contained: true
vignette: >
  %\VignetteIndexEntry{PACMOS-tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css, echo=FALSE}
pre,
pre.output,
pre.message,
pre.warning,
pre.sourceCode,
div.sourceCode {
  max-height: 250px;
  overflow-y: auto;
  overflow-x: auto;
}

```

# Overview

This vignette shows how to:

1. Add a query sample to MOFA reference inputs,
2. Run MOFA2,
3. Project query sample into reference MOFA space,
4. Infer archetype weights of query sample,
5. Plot query sample onto Pareto archetype space.

## Step 0: Set Up Python Environment

Before using MESOMICS, make sure you have the required Python environment.

Open your terminal and run:

```{bash, eval=FALSE}
conda env create -f mofa_env.yml
conda activate mofa_env
which python
# For example: /home/user/miniconda3/envs/mofa_env/bin/python
```


```{r setup}
library(PACMOS)
```

Example data paths
```{r setup-paths}
mofa_dir <- system.file("extdata/MESOMICS_references", package = "PACMOS")
reference_LFs  <- system.file("extdata/MESOMICS_references", "MESOMICS_latent_factors.csv", package = "PACMOS")
query_csv <- system.file("extdata/test_data", "test-normalised-gene_count.csv", package = "PACMOS")


out_dir <- file.path('.', "query_output")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
out_dir

Sys.setenv(MOFA_PYTHON = "/home/lipikal/miniconda3/envs/mofa_env/bin/python") #replace this with your 'which python' output
python_path <- Sys.getenv("MOFA_PYTHON", unset = NA)
```

## Step 1: Add query samples to MOFA inputs
This function integrates one or more query samples into reference multi-omic MOFA input matrices using an incremental approach. For each query sample, it aligns gene identifiers with the reference matrices, inserts observed values into the corresponding data layer, and fills all other layers with NA to maintain a consistent structure across views.

The function creates a structured output directory where each query sample has its own folder, containing an inputs/ subdirectory with all .RData matrices required for downstream MOFA training.

```{r run-step1, results='markup', warning=FALSE}

## -----------------------------------------------------------------------------
print('Starting Step 1')
PACMOS::s1_add_sample_to_mofa(
      query_matrix_path = query_csv,
      mofa_dir = mofa_dir,
      value_data_types = c('D_exprB_MOFA'),
      outdir = out_dir,
      python_bin = python_path
    )
cat("Outputs are saved in:", out_dir, "\n")

```

## Step 2: Train MOFA models
This function trains MOFA models for each query sample using the prepared input matrices. It loads the multi-omic inputs generated in Step 1, configures the model according to the specified views and data types (including binary layers), and runs MOFA.

Each query sample is processed independently, producing a trained MOFA model.

```{r run-step2, warning=FALSE}
print('Starting Step 2')

PACMOS::s2_run_mofa(
  models_dir      = out_dir,
  matrices_subdir = "inputs",
  num_factors = 10,
  convergence_mode = "slow",
  maxiter = 10000,
  use_basilisk = FALSE,
  skip_existing = TRUE,
  python_bin      = python_path,
  views_map = c(
    RNA     = "D_exprB_MOFA",
    MethPro = "D_met.proB_MOFA",
    MethBod = "D_met.bodB_MOFA",
    MethEnh = "D_met.enhB_MOFA",
    Total   = "D_cnv_MOFA",
    Minor   = "D_loh_MOFA",
    Alt     = "D_alt_MOFA"
  ),
  binary_views = "Alt"
)
cat("Outputs are saved in:", out_dir, "\n")

```

## Step 3: Visualize query samples
This function projects query samples into the reference latent factor space and generates quality control visualizations. It compares the learned latent factors of retrained models with reference axes, computes correlations, and produces diagnostic plots to assess how well the query samples align with known biological structure.

The outputs are saved as PDF files for each sample in the `plots` directory inside `inputs`.

```{r run-step3,warning=FALSE}
print('Starting Step 3')

sample_ids <- list.files(out_dir)

for (id in sample_ids) {
  PACMOS::s3_plot_query_samples_mofa(
    models_dir      = out_dir,
    matrices_subdir = "inputs",
    query_sample    = id,
    reference_LFs   = reference_LFs,
    reference_axes  = c(
      "Morphology_LF",
      "Adaptive-response_LF"
    ),
    group = "group1",
    python_bin = python_path
  )
}
cat("Outputs are saved in:", out_dir, "\n")

```

## Step 4: Infer archetype weights
This function estimates the contribution of predefined biological archetypes to each query sample using a fuzzy weighting approach. Based on the latent factor coordinates obtained from MOFA, it calculates the degree to which each sample belongs to each archetype.

The result is saved as a csv file in the `plots` directory.

```{r run-step4, warning=FALSE}
print('Starting Step 4')

archetype_coords <- data.frame(
  Archetype = c(
    "Cell division",
    "Tumor-immune-interaction",
    "Acinar"
  ),
  Morphology_factor.MESOMICS = c(
    -3.59733245242571,
    -1.79807036681759,
    3.85856701073313
  ),
  Adaptive.response_factor.MESOMICS = c(
    -2.39600999565261,
    3.56348237698459,
    -0.825586204203109
  ),
  stringsAsFactors = FALSE
)

PACMOS::infer_fuzzy_weights(
  models_dir      = out_dir,
  matrices_subdir = "inputs",
  coord           = archetype_coords,
  n_archetypes    = 3
)
cat("Outputs are saved in:", out_dir, "\n")

```

## Step 5: Plot results
This function visualizes the archetype composition of query samples based on the inferred fuzzy weights.

The plot is saved as a pdf file in the `plots` directory.

```{r run-step5, warning=FALSE}
print('Starting Step 5')

PACMOS::plot_fuzzy_query_sample(
  models_dir      = out_dir,
  matrices_subdir = "inputs"
)
cat("Outputs are saved in:", out_dir, "\n")

```

